Resolvable Ambiguity

### 2019-07-04 paper/paper.tex
Before splitting work

\newcommand{\support}{\mathit{Supp}}
\newcommand{\NT}{V} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{L} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}
\newcommand{\reqpl}{(}
\newcommand{\reqpr}{)}
\newcommand{\reqp}[1]{\reqpl#1\reqpr}
\newcommand{\pospl}{[}
\newcommand{\pospr}{]}
\newcommand{\posp}[1]{\pospl#1\pospr}

@@

Outline of introduction, one bullet per paragraph, roughly:

\begin{itemize}
\item Creating languages, particularly DSLs, it's nice to create pieces and compose. Composing syntax in particular quickly creates ambiguous grammars. Two broad approaches: forbid/prevent ambiguities at all (statically), or handle at parse-time (dynamically). Former more well-explored than the latter, which we focus on. The former is essentially the ambiguity problem, but the latter has not been formally defined. (To my knowledge we're first in presenting resolutions as opposed to interpretations (ASTs), which means that what we're talking about here is a little bit different from what previous approaches have done)

\item Motivated by issues in \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011} ''we introduce a formal definition for resolvable ambiguity.'' (They both merely present ASTs, which are unlikely to be nice for an end-user. Probably want better motivation than ''nice.'') \ldots It turns out that this is not always possible, prompting an additional problem: can we statically determine if a language is resolvably ambiguous? We call the former the dynamic resolvability problem, and the latter the static resolvability problem.

\item The dynamic problem is ''easy'' (though we haven't found any previously presented solution), but the static problem is not. We present a subclass of languages for which we have a decidable, sound, and complete decision procedure for the static problem, but also two larger classes of languages for which the algorithm is merely decidable and sound, or not applicable at all, respectively.
\end{itemize}

\noindent\hrulefill

When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments. However, how to make such compositions well-behaved is not always obvious. In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous. Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous words. i.e., ambiguous programs.

\noindent\hrulefill

% I really don't like this paragraph. I'm trying to say that much work has been done on ambiguity, and that work in other fields include it, while PL tends to be more narrow-minded.
Formal languages have long had the concept of ambiguity, of words that can be constructed in multiple ways, for example using multiple distinct derivations of a context-free grammar. In linguistics, this is important, since human languages tend to be highly ambiguous, and as such much work has been done on the subject. Examples include general parsing algorithms \cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}, ambiguity detection (most importantly its undecidability \cite{cantorAmbiguityProblemBackus1962}, but also various pragmatic approaches \cite{brabrandAnalyzingAmbiguityContextFree2007,axelssonAnalyzingContextFreeGrammars2008,bastenAmbiguityDetectionProgramming2011})

However, when using these formal methods to specify the syntax for programming languages, ambiguity becomes a strictly undesirable property. We tend to want to give the compiler a single interpretation of any given program, and so we work hard to make programming language grammars entirely unambiguous. This view, that eliminating ambiguity is \emph{the} correct thing to do, is highly prevalent (e.g. \cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).

Some exceptions do exist however. Closest to our approach are \citet{danielssonParsingMixfixOperators2011}, who ''feel that it is overly restrictive to require the grammar to be unambiguous'', and suggest that only ambiguous \emph{parses} should be rejected, not ambiguous grammars. This seems to stem largely from the fact that their grammars are mostly constructed from user-defined operators, and so cannot be expected to compose without ambiguity. Instead, in case of an unambiguous parse, they present an error that includes the possible parses, to assist in debugging.

We propose that this idea---treating ambiguity as a parse-time error---can be beneficial even when the entire grammar is constructed by a single language designer, if done in a more rigorous fashion. Consider the following nested match expression in OCaml (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:

@@

\setlength{\grammarindent}{9.5em}
\begin{grammar}
<expr> ::= 'match' <expr> 'with' <pattern-matching>

<pattern-matching> ::= '|'? <pattern> ('when' <expr>)? \\
('|' <pattern> ('when' <expr>)? '->' <expr>)*
\end{grammar}

\noindent If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match. The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended. Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user. Instead, we can present the alternatives in the form of modified code that parses unambiguously as the corresponding alternative:

@@

\noindent Unfortunately, not all ambiguities can be resolved in this way. Again, looking at an extract from the informal OCaml grammar:

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <expr> ';' <expr>
  \alt '[' <expr> (';' <expr>)* ';'? ']'
  \alt <constant>
\end{grammar}

\noindent The first production is sequential composition, the second is lists (the empty list is under \synt{constant}). Now consider the expression ''\ocaml{[1; 2]}''. We find that it is ambiguous with two alternatives:
\begin{enumerate}
  \item A list with two elements.
  \item A list with one element, namely a sequential composition.
\end{enumerate}

\noindent We can write the latter option unambiguously by putting parentheses around ''\ocaml{1; 2}'', but there is no way to unambiguously write the former. OCaml itself solves this by unambiguously parsing the expression as a two-element list. In this case, the ''arbitrary'' choice seems reasonable; there is no other way to write a two-element list, ambiguously or not\footnote{Ignoring the cons operator, since this syntax is supposed to be syntactic sugar for it.}, hence we would rather like ''\ocaml{[1; 2]}'' to be interpreted as such.

The possibility of an unresolvable ambiguity is troubling however, since it represents an error that a user might encounter but only a language designer can solve; it requires a change to the grammar. As such, we propose to allow ambiguous programming language grammars, as long as all possible ambiguities can be resolved, and only require \emph{programs} to be unambigous. We call this property \emph{resolvable ambiguity}.

We claim the following contributions:

\begin{itemize}
\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp. word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
  \begin{itemize}
  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound for the first two versions, and complete for the first (Section~\ref{sec:static}).
  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
  \end{itemize}
\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
  \begin{itemize}
  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
  \end{itemize}
\end{itemize}

@@

A \emph{bag} (alternatively known as a \emph{multiset}) is a generalization of a set, it is an unordered collection where each element may appear multiple times. The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$. A bag $A$ is included in another bag $B$, written $A \subseteq B$, iff $\forall x.\ m_A(x) \leq m_B(x)$. The underlying set of elements of a bag $B$ is called the \emph{support} of $B$, and is given by $\support(B) = \{ x \mid m_B(x) > 0 \}$.

@@

$$
\begin{array}{r@{\ =\ }l}
  L(t) & \{a\} \\
  L(\epsilon) & \{\epsilon\} \\
  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \\
  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
\end{array}
$$

\noindent We will denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.

@@

A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is Kleene-star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.

@@

The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{left-most derivations}. A left-most derivation is a derivation where the non-terminal being replaced is always the left-most one.

\begin{definition}
A word $w \in L(G)$ is ambiguous if there are two distinct left-most derivations of that word.
\end{definition}

@@

A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols. Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states. $\lambda$ essentially means ''ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).

@@

\begin{itemize}
\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$.
\item $p \xrightarrow{i} q$, where $i \in \T_i$.
\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$.
\end{itemize}

@@

The partition names stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}. We will instead primarily use $\T_c$ and $\T_r$ for balanced parentheses.

@@

\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run, none are redundant. Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
\end{itemize}

@@

Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense. We will call this meaning an \emph{interpretation} of a word. Note that a single word may have multiple interpretations. As such, we will define a language as:

@@

\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees. In such a language, $\parse(1 + 2 * 3)$ would produce a set containing the two ASTs in Figure~\ref{fig:arith-example}:

\begin{figure}[t]
\begin{tabular}{cc}
  \Tree [.{$+$}
    1
    [.{$*$}
      2
      3 ] ] &
  \Tree [.{$*$}
    [.{$+$}
      1
      2 ]
    3 ] \\
\end{tabular}
\caption{The two allowable interpretations of $1 + 2 * 3$ in an arithmetic language without precedence.}
\label{fig:arith-example}
\end{figure}

@@

\begin{itemize}
\item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ that relates words to their left-most derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

@@

\noindent The example in Figure~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 * 3)$ and $(1 + 2) * 3$ are unambiguously interpreted as the first and second tree, respectively.

@@

We thus propose to loosen the common ''no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous. However, merely having an arbitrary function $\parse$ gives us very little to work with, and no way to decide whether the language it defines is resolvably ambiguous or not. The remainder of this paper will thus consider $\parse$ functions defined with a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, that gives us some decidable properties.

@@

\begin{definition}
  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unresolvable}$) solving the static resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse) = \mathit{resolvable}$ implies that the language given by $\parse$ is resolvably ambiguous, and $f(\parse) = \mathit{unresolvable}$ implies that the language given by $\parse$ is not resolvably ambiguous.
  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$.
  \end{itemize}
  \label{def:static-procedure}
\end{definition}

@@

\begin{definition}
  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unresolvable}$) solving the dynamic resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
  \end{itemize}
  \label{def:dynamic-procedure}
\end{definition}

@@

With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}. Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label. Also note that the right-hand side of a production is here a regular expression, rather than the theoretically simpler sequence used in a context-free grammar. For proof technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense. This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}. In practice, this is likely desireable regardless, since a user is likely to attach different semantic meaning to each occurrence of a non-terminal.

Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal. To lessen clutter, we will write $E_\emptyset$ as $E$. As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example. In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication. By ''direct child'' we mean ''without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.

@@

\begin{itemize}
\item $G_D$ represents a word language describing all semantically distinct programs.
\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation (i.e., the result of marks).
\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
\end{itemize}

>>>>

\newcommand{\support}{\mathit{Supp}}
\newcommand{\NT}{V} % Set of nonterminals
\newcommand{\T}{\Sigma} % Set of terminals
\newcommand{\Labels}{L} % Set of labels
\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
\newcommand{\range}[2]{#1\!-\!#2}
\newcommand{\regex}{\mathit{Reg}}
\newcommand{\reqpl}{(}
\newcommand{\reqpr}{)}
\newcommand{\reqp}[1]{\reqpl#1\reqpr}
\newcommand{\pospl}{[}
\newcommand{\pospr}{]}
\newcommand{\posp}[1]{\pospl#1\pospr}
\newcommand{\dfa}{\mathit{dfa}} % Go from label to its corresponding dfa
\newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
\newcommand{\nt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs

@@

When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments. However, how to make such compositions well-behaved is not always obvious. In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous. Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs. The former is well explored in the form of heuristics for ambiguity detection (e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny. This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g. \cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).

Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g. \cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}. However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity. Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation. For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.

In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists. For example, in a language with semicolon for both sequential composition and element separation in a list (e.g. OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$. The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative. This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}. This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.

% TODO: I feel like there should be a clear transition between the more pracctical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.

% TODO: resolvable ambiguity cannot be defined solely in terms of a grammar, we require interpretations of words as well, thus we have to introduce a new formalism that does this, the formalism is an intuitive extension of EBNF grammars. this is probably written as a contrast with classical ambiguity, which should be described briefly for contrast

Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)? For classical ambiguity in context-free grammars, the former problem is solved by any general parser, while the latter is undecidable \cite{cantorAmbiguityProblemBackus1962}. For resolvable ambiguity, in a formalism that associates trees with EBNF-grammars, the dynamic problem is similarly solvable, and the static problem is similarly difficult. We do not yet know if the latter is undecidable in general, but we have found a subclass of languages for which it is decidable.

More specifically, this paper makes the following contributions:

\begin{itemize}
\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp. word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
  \begin{itemize}
  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
  \end{itemize}
\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
  \begin{itemize}
  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
  \end{itemize}
  % Briefly say something about the tool, that it lies ``on top of'' what's presented herein
\end{itemize}

\section{Motivating Example}

This section walks through an example from OCaml where an ambiguity error can be clearer than what the actual compiler produces, suggesting that (resolvable) ambiguity has merit even in the context of a general purpose programming language.

Consider the following nested match expression (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:

@@

\setlength{\grammarindent}{9.5em}
\begin{grammar}
<expr> ::= \verb|'match'| <expr> \verb|'with'| <pattern-matching>

<pattern-matching> ::= \verb{'|'{$^{?}$ <pattern> (\verb|'when'| <expr>)$^{?}$ \\
(\verb{'|'{ <pattern> (\verb|'when'| <expr>)$^{?}$ \verb|'->'| <expr>)*
\end{grammar}

\noindent If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match. The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended. Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user. Instead, we can present the alternatives in the form of modified code that parses unambiguously as each corresponding alternative:

@@

@@

A \emph{bag}, or \emph{multiset}, is a generalization of a set; it is an unordered collection where each element may appear multiple times. The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$. A bag $A$ is included in another bag $B$, written $A \subseteq B$, iff $\forall x.\ m_A(x) \leq m_B(x)$. The underlying set of elements of a bag $B$ is called the \emph{support} of $B$, and is given by $\support(B) = \{ x \mid m_B(x) > 0 \}$.

@@

$$
\begin{array}{r@{\ =\ }l}
  L(t) & \{a\} \\
  L(\epsilon) & \{\epsilon\} \\
  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \} \\
  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
\end{array}
$$

\noindent We denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.

@@

A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is the Kleene star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.

@@

The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{leftmost derivations}. A leftmost derivation is a derivation where the non-terminal being replaced is always the leftmost one.

\begin{definition}
A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of that word.
\end{definition}

@@

A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols. Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states. $\lambda$ essentially means ``ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).

@@

\begin{itemize}
\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$; or
\item $p \xrightarrow{i} q$, where $i \in \T_i$; or
\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$,
\end{itemize}

@@

The names of the partitions stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.

@@

\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant. Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
\end{itemize}

@@

Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense. We will call this meaning an \emph{interpretation} of a word. Note that a single word may have multiple interpretations. As such, we define a language as:

@@

\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees. In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs in Fig.~\ref{fig:arith-example}:

\begin{figure}[t]
\begin{tabular}{cc}
  \Tree [.{$+$}
    1
    [.{$\cdot$}
      2
      3 ] ] &
  \Tree [.{$\cdot$}
    [.{$+$}
      1
      2 ]
    3 ] \\
\end{tabular}
\caption{The two allowable interpretations of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
\label{fig:arith-example}
\end{figure}

@@

\begin{itemize}
\item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

@@

\noindent The example in Fig.~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.

@@

We thus propose to loosen the common ``no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous. However, merely having an arbitrary function $\parse$ gives us very little to work with, and no way to decide whether the language it defines is resolvably ambiguous or not. The remainder of this paper will thus consider $\parse$ functions defined using a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, which gives us some decidable properties.

@@

\begin{definition}
  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the static resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse) = \mathit{resolvable}$ implies that the language given by $\parse$ is resolvably ambiguous, and $f(\parse) = \mathit{unresolvable}$ implies that the language given by $\parse$ is not resolvably ambiguous.
  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$.
  \end{itemize}
  \label{def:static-procedure}
\end{definition}

@@

\begin{definition}
  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
  \begin{itemize}
  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
  \end{itemize}
  \label{def:dynamic-procedure}
\end{definition}

@@

With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Fig.~\ref{fig:input-language-definition}. Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label. Also note that the right-hand side of a production is a regular expression, rather than the theoretically simpler sequence used in a context-free grammar. For proof-technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense, when marks are removed. This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}. In practice, this is likely desireable regardless, since a user is likely to attach different semantic meaning to each occurrence of a non-terminal.

Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal. To lessen clutter, we will write $E_\emptyset$ as $E$. As an example, consider the language definition in Fig.~\ref{fig:running-example-definition}, which will be used as a running example. In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication. By ``direct child'' we mean ``without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.

@@

\begin{itemize}
\item $G_D$ represents a word language describing all semantically distinct programs.
\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding grouping parentheses and other forms of disambiguation (i.e., the result of marks).
\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
\end{itemize}

\noindent Fig.~\ref{fig:running-example-generated} shows the four grammars generated from our running example in Fig.~\ref{fig:running-example-definition}. The context-free grammars are produced by a rather standard translation from regular expressions to CFGs, while the primed grammars get a new non-terminal per distinctly marked non-terminal in $D$, where each new non-terminal only has the productions whose label is not in the mark. For example, the non-terminal $E_{\{a\}}$ in Fig.~\ref{fig:running-example-generated:t-prime} has no production corresponding to the $a$ production in Fig.~\ref{fig:running-example-definition}.
